Tree establishment in models of forest dynamics: a key priority for further research

### 2023-04-25 manuscript/manuscriptRW.Rmd
update references table 1

---
title: "Tree establishment in models of forest dynamics: a key priority for further research"
author:
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Tim Anders,
- Kristin H. Braziunas,
- Josef	Bruna,
- Samuel Fischer,
- Jessica Hetzer,
- Thomas Hickler,
- Christian	Hochauer,
- Manfred J. Lexer,
- Heike	Lischke,
- Mats Mahnken,
- Paola	Mairota,
- Katarina Merganicova,
- Tobias Mette,
- Xavier Morin,
- Werner Rammer,
- Christopher P.O. Reyer,
- Simon	Scheiter,
- Daniel Scherrer,
- Harald Bugmann
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: yes
    number_sections: yes
    toc: no
    extra_dependencies: "subfig"
  word_document:
    reference_docx: template_paper.docx
mainfont: Arial
fontsize: 12pt
linestretch: 1.5
bibliography: Regeneration.bib
link-citations: yes
nocite: |
  @bugmann1996a, @bohn2014, @fabrika2005, @mette2014, @smith2001a, @langan2017, @seidl2012a, @lischke2006, @schumacher2006, @scheiter2009, @scheller2004, @morin2021, @lasch-born2020, @hickler2012, @fischer_lessons_2016
linkcolor: blue
csl: global-change-biology.csl
editor_options:
  markdown:
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

Most of the models overestimate ingrowth levels, which is only
compensated by some models with high simulated mortality rates in the
early stages of population dynamics. Simulated species diversity of
ingrowth matches the observed ranges quite well. Models simulating
higher species diversity at the stand level do not feature higher
ingrowth diversity. The effect of light availability on ingrowth levels
is captured better than the effect of temperature and soil moisture, but
patterns are not consistent across models. The ingrowth of *Abies alba*
is captured well by most models across temperature and soil moisture
gradients, whereas ingrowth of *Fagus sylvatica* is generally
underestimated. *Picea abies* ingrowth is generally overestimated,
particularly at colder sites. Increasing complexity in the tree
establishment modules of the models is not related to higher accuracy of
simulated tree establishment. Furthermore, individual model design is
more important than scale (stand, landscape, global) and approach
(empirical, process-based) to accurately capture tree establishment.

Despite considerable mismatches between simulation results and data in
some instances, it is remarkable that most of the models capture the
essential features of the highly complex process of tree establishment,
while not having been parameterized with such data. We conclude that in
spite of all the uncertainties, a lot can be gained by a focus on the
modeling of establishment processes in these models. This has the
potential to render long-term projections of forest dynamics under
changing environmental conditions much more robust.

@@

Forests provide a wide range of products and services of vital
importance to humankind [@foresteurope2020]. Under the pressure of
climate change, increasing disturbances and changing societal demands on
forests ecosystems services, it is becoming ever more important to
understand how forest structure, composition and function will change,
and to evaluate forest capacity to adapt to or be resilient after
disturbance [@lindner2010; @seidl2022]. A wide range of models of forest
dynamics were developed over the past decades considering the impacts of
climate on forest dynamics [@bugmann2022]. From these studies, it is
evident that we have a reasonably good understanding how to model tree
growth [@vanclay1997; @bugmann1996], and substantial efforts have been
dedicated to improving the representation of tree mortality
[@bugmann2019; @cailleret2017]. In contrast, tree establishment is a
much less studied process that is often represented rather coarsely in
these models [@walck2011; @leishman1992; @price2001], which is a clear
research gap in the context of climate-induced forest disturbances and
forest resilience.

Tree establishment arises from multiple processes such as pollination,
fruit maturation, seed production, dispersal, germination, juvenile
growth and survival [@price2001; @vacchiano2018]. All these processes
are difficult to assess, and some of them are scarcely understood and
thus appear highly stochastic [@bogdziewicz2021]. Identifying the
appropriate level of complexity for the mathematical formulation of the
key factors that are leading to successful tree establishment is
challenging. Currently, tree establishment processes in dynamic forest
models are handled in a multitude of ways [@konig_tree_2022;
@bugmann2022]: from a) entirely ignoring it [as done in classical forest
growth models, e.g., Pretzsch et al. -@pretzsch2002], across b) the use
of a few simple environmental filters, as done in most forest "gap"
models [@shuman2011] and Dynamic Global Vegetation Models [e.g.,
@hickler2012; @smith2001], to c) complex approaches that incorporate
local feedback from the canopy, multiple ecological processes and often
also small time steps [e.g., @seidl2012; @wehrli2006], or d) field-based
statistical parameterizations, which however are not easy to extrapolate
in space and time [e.g., @ribbens1994].

Overall, models are needed to (1) synthesize existing empirical data and
explore their consequences, (2) assess future tree establishment e.g. in
the context of global change scenarios, and (3) identify the most
important processes that are shaping ecological patterns. Given the
current strategies that are used in models of forest dynamics to
represent tree establishment, model behavior often is prone to problems,
such as very high levels of tree establishment that necessitate excess
mortality at early stages of tree life to simulate correct stand
structure and composition. Also, inadequately high species diversity in
tree establishment may be simulated, which is characteristic of
"classical" forest gap models [@gutiérrez2016], at least as long as the
simulations comprise a multitude of species. Some models use calibration
against local canopy-level data to define appropriate ingrowth levels,
which is likely to strongly hamper the general applicability of these
models e.g. under scenarios of climate change. Furthermore, correctly
capturing the species composition in tree establishment as a function of
the presence of seed trees in the canopy is often a particular
challenge, potentially leading to unrealistic successional drift in the
model, which must be corrected by factors that are hard to parameterize
[e.g., @lischke2006]. A related issue is the excessive reduction of
species diversity due to positive feedback effects such that eventually
just single-species stands remain. This is sometimes corrected by the
incorporation of a low level of seed influx of all species at all times
[@schumacher2006] or by restricting the number of seeds per species in
the seedbank [@lischke2006]. However, simulated species composition is
usually exceedingly sensitive to such a formulations, whereas the
parameters of such functions are poorly constrained by field data.
Lastly, there is often a problem with insufficient observational
constraints on parameter values for models that start from very small
tree sizes (e.g., 10 cm of height or even seeds) and track tree
development in a process-oriented manner by considering a multitude of
ecological influences, rather than emphasizing tree establishment into a
larger size class [e.g., @wehrli2006].

Thus, a focus on the modeling of tree establishment processes is sorely
needed and overdue [@price2001; @walck2011] if we are to make reliable
projections of future forest dynamics, i.e., when the models need to be
operated in extrapolation mode, as well as from a fundamental ecological
point of view for increasing systems understanding. Also in empirical
ecology, tree establishment has not received a lot of attention
[@hanbury-brown2022], although many sub-processes have been studied in
detail. Yet, little data is available that covers all the processes
within one species along environmental gradients, let alone for a vast
suite of species. Thus, the available data are fragmented, which
constitutes a major problem for model building [@clark1999]. On the one
hand, monitoring on permanent sample plots (such as National Forest
Inventories) often measures tree establishment [@zell2019], however with
a design that captures the rate after the stand initiation phase
[@hallsby2015]. Data from permanent plots, although available across
regions, are highly heterogeneous, rendering their use difficult in a
modeling context [@käber2021]. On the other hand, targeted studies to
measure tree establishment on experimental sites are often limited in
spatial extent, species studied, or the sub-set of processes that are
targeted [@berdanier2016; @collet2006].

In the present study, for the first time, a large number of forest
models commonly used to assess forest dynamics under climate change are
evaluated against a continental-scale, multi-species harmonized dataset
on tree establishment [@käber2022]. By tree establishment, we refer to
the passing of trees across a specific diameter threshold ("ingrowth").
We included models that are based on a range of "philosophies", from
models operating at the stand to the global scale as well as the range
of models from empirically derived to "process-based" [e.g.,
@bugmann1996; @fabrika2012; @lexer2001; @reyer2014]. Because of the
large variability in tree establishment patterns in nature and the large
number of factors driving this process --- including some that are not
incorporated in most models such as deer browsing --- we do not aim at a
detailed statistical evaluation of each model. Instead, we aim at
evaluating the general ingrowth patterns and magnitudes simulated by the
models, and benchmark the simulated regeneration niche of multiple
species against empirical data along a wide environmental gradient of
temperature, moisture, and light availability.

@@

The establishment modules can be categorized as 'regeneration' or
'recruitment' models [@vanclay1994; @konig_tree_2022] (see Table
\@ref(tab:tableModels)). Regeneration models include processes such as
flowering and pollination, seed production, seed dispersal, germination,
and seedling growth, which ultimately lead to the simulated number of
established trees. Recruitment models, in contrast, introduce a certain
number of new trees with certain characteristics such as biomass or
diameter, without explicitly considering earlier development processes.
We can further distinguish models that feature a feedback in the
establishment, i.e., where the level and species composition of
establishment is influenced by the existing adult tree community via the
production of seeds, seedlings or saplings, from models that do not
contain such a feedback.

@@

Ingrowth data covering a wide range of environmental conditions are hard
to obtain, and this is one of the reasons why most models of forest
dynamics have never been confronted with a dataset covering such
gradients over a large number of sites, to evaluate how well ingrowth is
captured. The observations used here are derived from a novel and
unprecedented network of sites in forest reserves that represent the
range of environmental gradients in temperature and precipitation in
central Europe as compiled in the framework of the EuFoRIa network
[@euforia2019] (Figure \@ref(fig:map)). These forests have been
unmanaged for at least 10 years prior to their designation as reserves,
and most of them provide time series of natural forest dynamics over
multiple decades and up to ca. 80 years. The census periods range from 3
to 37 years, with an average of 14 years. The data provide information
at the tree level, thus allowing for sequential comparison such as
individual-tree ingrowth and death between the measurements.

We selected 200 sites from this network as the benchmarking dataset for
the simulation. The selection of sites is representative for the
environmental variation of the data. This was achieved by applying
k-means clustering to define 200 clusters of plots from the original set
of 869 plots along the environmental dimensions of temperature, climatic
water balance, soil quality and exposition (i.e., slope and aspect).
Each of these sites featured at least two consecutive measurements.
Ingrowth thresholds for the sites differed between diameters of 0 and 10
cm. For our study, we defined two datasets where one included 165 sites
with a diameter threshold of 7 cm or lower, and the other included
another 35 sites with diameter thresholds between 7 and 10 cm.

@@

The overarching goal of the experiments was to assess the tree
establishment as it arises from empirical data against its
representation in a wide range of models of forest dynamics. We define
the tree establishment as the passing of a breast height diameter
threshold of 7 or 10 cm, respectively (synonym: ingrowth). To this end,
each modeling group was provided with a detailed protocol
[@díaz-yáñez2022] with instructions how to perform the simulations,
providing input variables on climate and soil conditions, and the
expected output variables. Neither were further site information (except
for the data specified below) nor any data on tree ingrowth or forest
stand features provided prior to the simulation. That is, the models
were run in "blind flight" mode.

The input variables were collected and aggregated to be adapted to the
needs of the different models. Time series of climatic variables from
1981 to 2018 were provided in hourly [Era5-land data, Muñoz Sabater
-@muñozsabater2019], daily and monthly resolution [CHELSA data, Karger
et al. -@karger2021]. Some variables required for some models, such as
relative humidity or vapor pressure deficit, were calculated from these
variables. The final strategy how to use the climate data was different
in each model, based on the approach that best suited the model (Table
\@ref(tab:tableModels)). Soil quality data were provided as continuous
values between 1 and 5 [Soilgrids dataset, Hengl et al. -@hengl2017].
The protocol also provided the elevation, slope and aspect for each of
the 200 sites [@asterscienceteam2019], but no other spatial information
such as coordinates, with the exception of iLand and aDGVM2, which
required blurred coordinates to derive highly detailed soil data.

@@

The simulations were run for mixed species forests (not multiple
single-species simulations) using mixtures of eleven species or genera
for which ingrowth data of sufficient quality were available from
EuFoRIa: *Fagus sylvatica, Picea abies, Abies alba, Carpinus betulus,
Tilia cordata, Acer pseudoplatanus, Betula spp., Fraxinus excelsior,
Quercus spp., Alnus glutinosa, Pinus sylvestris.* The same set of
species was used at all 200 sites. Two models included their standard
set of species for the simulations, which is much larger (i.e., ForClim
1, ForClim 11, TreeMig). In three models, fewer than these eleven
species were simulated (4C, xComp, LPJ-GUESS) (Table
\@ref(tab:tableModels)). In 4C only *Fagus sylvatica*, *Picea abies*,
*Betula* spp., *Quercus* spp. and *Pinus sylvestris* are parameterised.
XComp simulations did not consider *Fraxinus excelsior* due to a
depreciated species parameterisation. In LPJ-GUESS, *Acer
pseudoplatanus* and *Alnus glutinosa* are not parameterized and
therefore these species could not be included in the simulation.
Finally, the model aDGVM2 does not simulate individual species. Rather,
community assembly processes and trait filtering generate plant
communities that are adjusted to the biotic and abiotic conditions, and
the simulated plants can be classified into ecological strategies based
on their trait values in a post-processing step.

Each of the models reported the ingrowth number by sampling 200 times
over a 10-year interval for each species and per ha for each of the 200
sites. Multiple samples per site were used to better understand the
simulated variation within each site. This was done using different
strategies, depending on the model, including (1) sampling simulated
data from the same 1 ha plot in the equilibrium over time, (2) sampling
several 1 ha plots from the simulated forest at one specific point in
time (in the equilibrium) or (3) a combination of (1) and (2) . This
resulted in 880,000 observations per model that simulated the 11 species
included in the protocol (200 sites, 200 samples per site, 11 species
and two diameter thresholds). For the models that simulated additional
species, their ingrowth rates were aggregated as "others".

@@

```{r map, echo=FALSE, fig.cap="The 200 sites of the EuFoRIa network used in the present study, the colors show the dominant tree species per site in the observed ingrowth at the 10 cm threshold.", out.width='80%'}
knitr::include_graphics(here::here("figures", "theMap.png"))
```

@@

The simulation results were analyzed regarding (1) ingrowth levels
(i.e., ingrowth number per unit time and space), (2) species diversity
of the ingrowth, (3) ingrowth mortality, (4) model performance and model
traits, and (5) ingrowth gradients on regeneration niches. We evaluated
the species diversity across models and in relation to the observed data
by calculating the Shannon index based on the relative proportion of the
species in terms of basal area. The Shannon index was calculated for the
establishment ($H_{R_{n}}$) (Equation \ref{eqpiShannonR}) and at the
stand level ($H_{S_{n,s}}$) (Equation \ref{eqpiShannon}). The higher the
value of the index, the higher the species diversity at a particular
site and sample.

```{=tex}
\begin{equation}
H_{R_{n}} = - \sum_{i = 1}^{s}{p_{R_{n,i}}\ln(p_{R_{n,i}})}\label{eqpiShannonR}
\end{equation}
```
```{=tex}
\begin{equation}
p_{R_{n,i}} = \frac{r_{BA_{n,i}}}{TotalR_{BA_{n}}} \label{eqpiShannonR2}
\end{equation}
```
```{=tex}
\begin{equation}
H_{S_{n}} = - \sum_{i = 1}^{s}{p_{S_{n,i}}\ln(p_{S_{n,i}})}\label{eqpiShannon}
\end{equation}
```
```{=tex}
\begin{equation}
p_{S_{n,i}} = \frac{BA_{n,i}}{Total_{BA_{n}}} \label{eqpiShannon2}
\end{equation}
```
where $s$ is the total number of species present and that have a basal
area larger than zero in sample $n$; $p_{n,r,i}$ is the proportion of
the species $i$ in sample $n$ calculated as the ingrowth basal area
($r_{BA_{n,i}}$) for that species $i$ relative to the total recruited
basal area ($TotalR_{BA_{n}}$) of the sample $n$; $p_{S_{n,i}}$ is the
proportion of the species $i$ calculated as the basal area of all trees
($BA_{n,i}$) of that species $i$ relative to the total basal area
($TotalR_{BA_{n}}$) of the sample $n$. The species diversity was not
assessed for the model aDGVM2 because it does not simulate individual
species.

@@

```{=tex}
\begin{equation}
Reineke = (7/10)^{(-1.605)} \label{eqReineke}
\end{equation}
```
We assessed model performance in relation to model traits focusing on
(1) model complexity as defined by @bugmann2022, (2) model type
(empirical or process based), (3) the presence of absence of a canopy
feedback for ingrowth, and (4) the scale of application of the model
(stand, landscape or global). We tested for significant differences
using two-sided t-tests by considering each of these model features
regarding the observed and simulated ingrowth levels and species
diversity, respectively.

@@

```{=tex}
\begin{equation}
\bar{R}_{BA_{share_{t,i}}} = \frac{\bar{r}_{BA_{t,i}}}{\sum_{i = 1}^{s}{\bar{r}_{{BA_{t,i}}}}} \label{eq:BAshare}
\end{equation}
```
where $s$ is the total number of species simulated, and $\bar{r}_{t,i}$
is the mean basal area ingrowth of the species $i$ per site $t$ across
the available samples at that site.

We removed 35 sites from the comparisons at a threshold of 7 cm
ingrowth, because field data did not include these measurements.
Furthermore, regarding the share in the ingrowth basal area per species
($\bar{R}_{BA_{share_{i}}}$), we categorized this as zero when the
ingrowth basal area of that species ($\bar{r}_{BA_{t,i}}$) was zero but
also when both the ingrowth basal area of that species
($\bar{r}_{BA_{t,i}}$) and the total ingrowth basal area (
$\sum_{i = 1}^{s}{\bar{r}_{BA_{t,i}}}$ ) equaled zero.

@@

The majority of the models overpredicted ingrowth diversity; only two of
them underestimated ingrowth diversity (4C and Landis-II) (Figure
\@ref(fig:combinedRichness)B and Figure S4). Both the models with
feedback and those without feedback from the adult trees to ingrowth
(via seed production) overestimate, underestimated and captured
reasonably well ingrowth species diversity compared to the observations.
Again, simulated diversity was considerably less variable than observed
diversity in the ingrowth.

>>>>

---
title: "Tree establishment in models of forest dynamics: a key priority for further research"
author:
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Tim Anders,
- Kristin H. Braziunas,
- Josef	Bruna,
- Samuel Fischer,
- Jessica Hetzer,
- Thomas Hickler,
- Christian	Hochauer,
- Manfred J. Lexer,
- Heike	Lischke,
- Mats Mahnken,
- Paola	Mairota,
- Katarina Merganicova,
- Tobias Mette,
- Xavier Morin,
- Werner Rammer,
- Christopher P.O. Reyer,
- Simon	Scheiter,
- Daniel Scherrer,
- Harald Bugmann
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: yes
    number_sections: yes
    toc: no
    extra_dependencies: "subfig"
  word_document:
    reference_docx: template_paper.docx
mainfont: Arial
fontsize: 12pt
linestretch: 1.5
bibliography: Regeneration.bib
link-citations: yes
nocite: |
  @bugmann1996a, @bohn2014, @fabrika2005, @mette2014, @smith2001a, @langan2017, @seidl2012a, @lischke2006, @schumacher2006, @scheiter2009, @scheller2007, @morin2021, @lasch-born2020, @hickler2012, @fischer_lessons_2016, @langan2017, @scheiter2013 
linkcolor: blue
csl: global-change-biology.csl
editor_options:
  markdown:
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

Most of the models overestimate ingrowth levels, which is only
compensated by some models with high simulated mortality rates in the
early stages of individual trees dynamics. Simulated species diversity
of ingrowth matches the observed ranges quite well. Models simulating
higher species diversity at the stand level do not feature higher
ingrowth diversity. The effect of light availability on ingrowth levels
is captured better than the effect of temperature and soil moisture, but
patterns are not consistent across models. The ingrowth of *Abies alba*
is captured well by most models across temperature and soil moisture
gradients, whereas ingrowth of *Fagus sylvatica* is generally
underestimated. *Picea abies* ingrowth is generally overestimated,
particularly at colder sites. Increasing complexity in the tree
establishment modules of the models is not related to higher accuracy of
simulated tree establishment. Furthermore, individual model design is
more important than scale (stand, landscape, global) and approach
(empirical, process-based) for accurately capturing tree establishment.

Despite considerable mismatches between simulation results and data in
some instances, it is remarkable that most of the models capture the
essential features of the highly complex process of tree establishment,
while not having been parameterized with such data. We conclude that in
spite of all the uncertainties, much can be gained by evaluating and
refining model establishment processes. This has the potential to render
long-term projections of forest dynamics under changing environmental
conditions much more robust.

@@

Forests provide a wide range of products and services of vital
importance to humankind [@foresteurope2020]. Under the pressure of
climate change, increasing disturbances impacts and changing societal
demands on forests ecosystems services, it is becoming ever more
important to understand how forest structure, composition and function
will change, and to evaluate forest capacity to adapt to or be resilient
after disturbance [@lindner2010; @seidl2022]. A wide range of models of
forest dynamics were developed over the past decades considering the
impacts of climate on forest dynamics [@bugmann2022]. From these
studies, it is evident that we have a reasonably good understanding how
to model tree growth [@vanclay1997; @bugmann1996], and substantial
efforts have been dedicated to improving the representation of tree
mortality [@bugmann2019; @cailleret2017]. In contrast, tree
establishment is a much less studied process that is often represented
rather coarsely in these models [@walck2011; @leishman1992; @price2001],
which is a clear research gap in the context of climate-induced forest
disturbances and forest resilience.

Tree establishment arises from multiple processes such as pollination,
fruit maturation, seed production, dispersal, germination, juvenile
growth and survival [@price2001; @vacchiano2018]. All these processes
are difficult to assess, and some of them are scarcely understood and
thus appear highly stochastic [@bogdziewicz2021]. Identifying the
appropriate level of complexity for the mathematical formulation of the
key factors that are leading to successful tree establishment is
challenging. Currently, tree establishment processes in dynamic forest
models are handled in a multitude of ways [@konig_tree_2022;
@bugmann2022]: from 1) entirely ignoring it [as done in classical forest
growth models, e.g., Pretzsch et al. -@pretzsch2002], across 2) the use
of a few simple environmental filters, as done in most forest "gap"
models [@shuman2011] and Dynamic Global Vegetation Models [e.g.,
@hickler2012; @smith2001], to 3) complex approaches that incorporate
local feedback from the canopy, multiple ecological processes and often
also short time steps [e.g., @seidl2012; @wehrli2006], or 4) field-based
statistical parameterizations, which however are not easy to extrapolate
in space and time [e.g., @ribbens1994].

Overall, models are needed to (1) synthesize existing empirical data and
explore their relationships, (2) assess future tree establishment e.g.
in the context of global change scenarios, and (3) identify the most
important processes that are shaping ecological patterns. Given the
current strategies that are used in models of forest dynamics to
represent tree establishment, model behavior often is prone to problems,
such as very high levels of tree establishment that necessitate excess
mortality at early stages of tree life to simulate correct stand
structure and composition. Also, inadequately high species diversity in
tree establishment may be simulated, which is characteristic of
"classical" forest gap models [@gutiérrez2016], at least as long as the
simulations set-up comprise a multitude of species. Some models use
calibration against local canopy-level data to define appropriate
ingrowth levels, which is likely to strongly hamper the general
applicability of these models e.g. under scenarios of climate change.
Furthermore, correctly capturing the species composition in tree
establishment as a function of the presence of seed trees in the canopy
is often a particular challenge, potentially leading to unrealistic
successional drift in the model, which must be corrected by factors that
are hard to parameterize [e.g., @lischke2006]. A related issue is the
excessive reduction of species diversity due to positive feedback
effects such that eventually just single-species stands remain. This is
sometimes corrected by the incorporation of a low level of seed influx
of all species at all times [@schumacher2006] or by restricting the
number of seeds per species in the seedbank [@lischke2006]. However,
simulated species composition is usually exceedingly sensitive to
assumptions about seed availability, whereas the parameters of such
functions are poorly constrained by field data. Lastly, there is often a
problem with insufficient observational constraints on parameter values
for models that start from very small tree sizes (e.g., 10 cm of height
or even seeds) and track tree development in a process-oriented manner
by considering a multitude of ecological influences, rather than
emphasizing tree establishment into a larger size class [e.g.,
@wehrli2006].

Thus, a focus on the modeling of tree establishment processes is sorely
needed and overdue [@price2001; @walck2011] if we are to make reliable
projections of future forest dynamics, i.e., when the models need to be
operated in extrapolation mode, as well as from a fundamental ecological
point of view for increasing systems understanding. Also in empirical
ecology, tree establishment has not received a lot of attention
[@hanbury-brown2022], although many sub-processes have been studied in
detail [@miina2006]. Yet, little data is available that covers all the
processes within one species along environmental gradients, let alone
for a vast suite of species. Thus, the available data are fragmented,
which constitutes a major problem for model building [@clark1999]. On
the one hand, monitoring on permanent sample plots (such as National
Forest Inventories) often measures tree establishment [@zell2019],
however with a design that captures the rate after the stand initiation
phase [@hallsby2015]. Data from permanent plots, although available
across regions, are highly heterogeneous, rendering their use difficult
in a modeling context [@käber2021]. On the other hand, targeted studies
to measure tree establishment on experimental sites are often limited in
spatial extent, species studied, or the sub-set of processes that are
targeted [@berdanier2016; @collet2006].

In the present study, for the first time, a large number of forest
models commonly used to assess forest dynamics under climate change are
evaluated against a continental-scale, multi-species harmonized dataset
on tree establishment [@käber2022]. By tree establishment, we refer to
the passing of trees across a specific diameter threshold ("ingrowth").
We included models that are based on a range of "philosophies", from
models operating at the stand to the global scale as well as the range
of models from empirically derived to "process-based" [e.g.,
@bugmann1996; @fabrika2012; @lexer2001; @reyer2014]. Because of the
large variability in tree establishment patterns in nature and the large
number of factors driving this process --- including some that are not
incorporated in most models such as deer browsing --- we do not aim at a
detailed statistical evaluation of each model. Instead, we aim at
evaluating the general ingrowth patterns and magnitudes simulated by the
models, and benchmark the simulated regeneration niche of multiple
species against empirical data along a wide environmental gradient of
temperature, moisture, and light availability [@grubb1977].

@@

The establishment modules can be categorized as 'regeneration' or
'recruitment' models [@vanclay1994; @konig_tree_2022] (see Table
\@ref(tab:tableModels)). Regeneration models include processes such as
flowering and pollination, seed production, seed dispersal, germination,
and seedling growth, which ultimately lead to the simulated number of
established trees. Recruitment models, in contrast, introduce a number
of new trees with certain characteristics such as biomass or diameter,
without explicitly considering earlier development processes. We can
further distinguish models that feature a feedback in the establishment,
i.e., where the level and species composition of establishment is
influenced by the existing adult tree community via the production of
seeds, seedlings or saplings, from models that do not contain such a
feedback.

@@

Ingrowth data covering a wide range of environmental conditions are hard
to obtain, and this is one of the reasons why most models of forest
dynamics have never been confronted with a dataset covering such
gradients over a large number of sites, to evaluate how well ingrowth is
captured. The observations used here are derived from a novel and
unprecedented network of sites in forest reserves that represent the
range of environmental gradients in temperature and precipitation in
Central Europe as compiled in the framework of the EuFoRIa network
[@euforia2019] (Figure \@ref(fig:map)). These forests have been
unmanaged for at least 10 years prior to their designation as reserves,
and most of them provide time series of natural forest dynamics over
multiple decades and up to ca. 80 years. The census periods range from 3
to 37 years, with an average of 14 years. The data provide information
at the tree level, thus allowing for sequential comparison such as
individual-tree ingrowth and death between the measurements.

We selected 200 sites from this network as the benchmarking dataset for
the simulation. The selection of sites is representative of the
environmental variation of the data. This was achieved by applying
k-means clustering to define 200 clusters of plots from the original set
of 869 plots along the environmental dimensions of temperature, climatic
water balance, soil quality and exposition (i.e., slope and aspect).
Each of these sites featured at least two consecutive measurements.
Ingrowth thresholds for the sites differed between diameters of 0 and 10
cm. For our study, we defined two datasets where one included 165 sites
with a diameter threshold of 7 cm or lower, and the other included
another 35 sites with diameter thresholds between 7 and 10 cm.

@@

The overarching goal of the experiments was to assess the tree
establishment as it arises from empirical data against its
representation in a wide range of models of forest dynamics. We define
the tree establishment as the passing of a breast height diameter
threshold of 7 or 10 cm, respectively (synonym: ingrowth). To this end,
each modeling group was provided with a detailed protocol
[@díaz-yáñez2022] with instructions for how to perform the simulations,
providing input variables on climate and soil conditions, and the list
of expected output variables. Neither were further site information
(except for the data specified below) nor any data on tree ingrowth or
forest stand features provided prior to the simulation. That is, the
models were run in "blind flight" mode.

The input variables were collected from different data sources and
aggregated to be adapted to the needs of the different models. Time
series of climatic variables from 1981 to 2018 were provided in hourly
[Era5-land data, Muñoz Sabater -@muñozsabater2019], daily and monthly
resolution [CHELSA data, Karger et al. -@karger2021]. Some variables
required for some models, such as relative humidity or vapor pressure
deficit, were calculated from these variables. The final instructions
for using the climate data was different in each model, based on the
approach that best suited the model (Table \@ref(tab:tableModels)). Soil
quality data were provided as continuous values between 1 and 5
[Soilgrids dataset, Hengl et al. -@hengl2017]. The protocol also
provided the elevation, slope and aspect for each of the 200 sites
[@asterscienceteam2019], but no other spatial information such as
coordinates, with the exception of iLand and aDGVM2, which required
blurred coordinates to derive highly detailed soil data.

@@

The simulations were run for mixed species forests (not multiple
single-species simulations) using mixtures of eleven species or genera
for which ingrowth data of sufficient quality were available from
EuFoRIa: *Fagus sylvatica L., Picea abies L., Abies alba Mill., Carpinus
betulus L., Tilia cordata Mill., Acer pseudoplatanus L., Betula spp. L.,
Fraxinus excelsior L., Quercus spp. L., Alnus glutinosa L., Pinus
sylvestris L.* The same set of species was used at all 200 sites. Two
models included their standard set of species for the simulations, which
is much larger (i.e., ForClim 1, ForClim 11, TreeMig). In three models,
fewer than these eleven species were simulated (4C, xComp, LPJ-GUESS)
(Table \@ref(tab:tableModels)). In 4C only *Fagus sylvatica*, *Picea
abies*, *Betula* spp., *Quercus* spp. and *Pinus sylvestris* are
parameterised. XComp simulations did not consider *Fraxinus excelsior*
due to a depreciated species parameterisation. In LPJ-GUESS, *Acer
pseudoplatanus* and *Alnus glutinosa* are not parameterized and
therefore these species could not be included in the simulation.
Finally, the model aDGVM2 does not simulate individual species. Rather,
community assembly processes and trait filtering generate plant
communities that are adjusted to the biotic and abiotic conditions, and
the simulated plants can be classified into ecological strategies based
on their trait values in a post-processing step.

Each of the models reported the ingrowth number by sampling 200 times in
a 10-year interval for each species and per ha for each of the 200
sites. Multiple samples per site were used to better understand the
simulated variation within each site. This was done using different
strategies, depending on the model, including (1) sampling simulated
data from the same 1 ha plot in the equilibrium over time, (2) sampling
several 1 ha plots from the simulated forest at one specific point in
time (in the equilibrium) or (3) a combination of (1) and (2) . This
resulted in 880,000 observations per model that simulated the 11 species
included in the protocol (200 sites, 200 samples per site, 11 species
and two diameter thresholds). For the models that simulated additional
species, their ingrowth rates were aggregated as "others".

@@

```{r map, echo=FALSE, fig.cap="The 200 sites of the EuFoRIa network (EuFoRIa, 2019) used in the present study, the colors show the dominant tree species per site in the observed ingrowth at the 10 cm threshold.", out.width='80%'}
knitr::include_graphics(here::here("figures", "theMap.png"))
```

@@

The simulation results were analyzed regarding (1) ingrowth levels
(i.e., ingrowth number per unit time and space), (2) species diversity
of the ingrowth, (3) ingrowth mortality, (4) model performance and model
traits, and (5) ingrowth gradients along regeneration niches. We
evaluated the species diversity across models and in relation to the
observed data by calculating the Shannon index based on the relative
proportion of the species in terms of basal area. The Shannon index was
calculated for the establishment ($H_{R_{n}}$) (Equation
\ref{eqpiShannonR}) and at the stand level ($H_{S_{n,s}}$) (Equation
\ref{eqpiShannon}). The higher the value of the index, the higher the
species diversity at a particular site and sample.

```{=tex}
\begin{equation}
H_{R_{n}} = - \sum_{i = 1}^{s}{p_{R_{n,i}}\ln(p_{R_{n,i}})}\label{eqpiShannonR}
\end{equation}
```
```{=tex}
\begin{equation}
p_{R_{n,i}} = \frac{r_{BA_{n,i}}}{TotalR_{BA_{n}}} \label{eqpiShannonR2}
\end{equation}
```
```{=tex}
\begin{equation}
H_{S_{n}} = - \sum_{i = 1}^{s}{p_{S_{n,i}}\ln(p_{S_{n,i}})}\label{eqpiShannon}
\end{equation}
```
```{=tex}
\begin{equation}
p_{S_{n,i}} = \frac{BA_{n,i}}{Total_{BA_{n}}} \label{eqpiShannon2}
\end{equation}
```
where $s$ is the total number of species present and that have a basal
area larger than zero in sample $n$; $p_{R_{n,i}}$ is the proportion of
the species $i$ in sample $n$ calculated as the ingrowth basal area
($r_{BA_{n,i}}$) for that species $i$ relative to the total recruited
basal area ($TotalR_{BA_{n}}$) of the sample $n$; $p_{S_{n,i}}$ is the
proportion of the species $i$ calculated as the basal area of all trees
($BA_{n,i}$) of that species $i$ relative to the total basal area
($TotalR_{BA_{n}}$) of the sample $n$. The species diversity was not
assessed for the model aDGVM2 because it does not simulate individual
species.

@@

```{=tex}
\begin{equation}
Reineke = \frac{7}{10}^{-1.605} \label{eqReineke}
\end{equation}
```
We assessed model performance in relation to model traits focusing on
(1) model complexity as defined by @bugmann2022, (2) model type
(empirical or process based), (3) the presence or absence of a canopy
feedback for ingrowth, and (4) the scale of application of the model
(stand, landscape or global). We tested for significant differences
using two-sided t-tests by considering each of these model features
regarding the observed and simulated ingrowth levels and species
diversity, respectively.

@@

```{=tex}
\begin{equation}
\bar{R}_{BA_{share_{t,i}}} = \frac{\bar{r}_{BA_{t,i}}}{\sum_{i = 1}^{s}{\bar{r}_{{BA_{t,i}}}}} \label{eq:BAshare}
\end{equation}
```
where $s$ is the total number of species simulated, and $\bar{r}_{t,i}$
is the mean basal area ingrowth of the species $i$ per site $t$ across
the available samples at that site. Furthermore, regarding the share in
the ingrowth basal area per species ($\bar{R}_{BA_{share_{i}}}$), we
categorized this as zero when the ingrowth basal area of that species
($\bar{r}_{BA_{t,i}}$) was zero but also when both the ingrowth basal
area of that species ($\bar{r}_{BA_{t,i}}$) and the total ingrowth basal
area ( $\sum_{i = 1}^{s}{\bar{r}_{BA_{t,i}}}$ ) equaled zero.

@@

The majority of the models overpredicted ingrowth diversity; as already
observed in figure \@ref(fig:H710), only two of them underestimated
ingrowth diversity (4C and Landis-II) (Figure
\@ref(fig:combinedRichness)B and Figure S4). Both the models with
feedback and those without feedback from the adult trees to ingrowth
(via seed production) overestimate, underestimated and captured
reasonably well ingrowth species diversity compared to the observations.
Again, simulated diversity was considerably less variable than observed
diversity in the ingrowth.
